# R markdown for documentation 

We have been writing code in Rscript and all the result of your code is printed at the console. From now on, we will learn to use a powerful interactive notebook interface to write code and display results. 

In addition, you can also use multiple languages including R, Python, and SQL in R-markdown. We will later introduce how to insert python code in r-markdown.


Next, we will write code in r-markdown to implement some small applications. In particular, we have learned the statement of if...else if... for branching; the for/while-loop for repetitive task. Today, we will use these statements to write the first computer game in R. 


## rock, paper, scissors
Let's write a program to play the "rock, paper, scissors" with the computer. At each round, the computer and you will choose one from "rock, paper, scissors". Then you compare the computer's and your choice to determine 1) you won 2) computer won 3) tie. Since there are three different outcomes, we need if...else if...else statement. 

## verison 1:
```{r}
you="rock"   
computer="paper" 

if (computer==you){
  print("this is a tie.")
  
} else if((computer=="rock" & you=="paper")|(computer=="paper" & you=="scissors")|(computer=="scissors" & you=="rock") ){
  
  print("You won!")
} else{
  
  print("Oops, computer won!")
}

```

While version 1 is a boring game because you have already know the outcome of the game. In order not to have the pre-determined outcome, we need to let you and computer randomly choose from "rock, paper, scissors". There is a convenient function in R for random sampling: 

sample(vector, n, replace=TRUE) samples n elements from vector with replacement. With replacement means that the sampled element is put back for the next sampling. E.g., 
```{r}
sample(c("rock","paper","scissors"), 4, replace=TRUE)
```


## verison 2
With the sample() function, we can improve the program as below:

```{r, eval=FALSE}
#you=sample(c("rock","paper","scissors"), 1, replace=TRUE)

you=readline("please enter your action: rock, paper, scissors > ")
computer=sample(c("rock","paper","scissors"), 1, replace=TRUE)

print(paste("your action is", you))
print(paste("computer action is", computer))

if (computer==you){
  print("this is a tie.")
  
} else if((computer=="rock" & you=="paper")|(computer=="paper" & you=="scissors")|(computer=="scissors" & you=="rock") ){
  
  print("You won!")
} else{
  
  print("Oops, computer won!")
}

```

## Version 3
Let's revisit the "rock, paper, scissors" game. This time, you and computer will play "rock, paper, scissors" for three rounds. Whoever won more rounds will be the winner.

Playing 3 rounds is a repetitive task, we can use for-loop. Since the game will be playing multiple rounds, we will need to implement a score board to keep track of the score at each round (think of the score board in a basketball game).
```{r, eval=FALSE}
you_score=0
computer_score=0

for (i in 1:3){
  
  #you=sample(c("rock","paper","scissors"), 1, replace=TRUE)
  you=readline("please enter your action: rock, paper, scissors > ")
  computer=sample(c("rock","paper","scissors"), 1, replace=TRUE)

  print(paste("your action is", you))
  print(paste("computer action is", computer))
  
  if (computer==you){
    print("this is a tied round.")
  
  } else if((computer=="rock" & you=="paper")|(computer=="paper" &   you=="scissors")|(computer=="scissors" & you=="rock") ){
  
    print("You won in this round!")
    you_score=you_score+1
    
  } else{
  
    print("Oops, computer won in this round!")
    computer_score=computer_score+1
    
  }
}

# determine who won finally
print("The final score is: ")
print(paste("you: ", you_score, " V.S. ", "computer: ", computer_score, sep=""))

if (you_score>computer_score){
  print("You won!")
} else if (you_score==computer_score){
  print("tie game")
} else {
  print("Computer won!")
}

```

## verision 4
Define a rps function for re-use:
```{r, eval=FALSE}
# determine the winner based on action
rps=function(you, computer){
  computer_score=0
  you_score=0
  
  print(paste("your action is", you))
  print(paste("computer action is", computer))
  
  if (computer==you){
    print("this is a tied round.")
  
  } else if((computer=="rock" & you=="paper")|(computer=="paper" &        you=="scissors")|(computer=="scissors" & you=="rock") ){
  
    print("You won in this round!")
    you_score=you_score+1
    
  } else{
    print("Oops, computer won in this round!")
    computer_score=computer_score+1
  }
  
  return(c(computer_score,you_score))
}

# the function to print the score 
winner=function(computer_score,you_score){
  print("The final score is: ")
  print(paste("you: ", you_score, " V.S. ", "computer: ", computer_score, sep=""))

  if (you_score>computer_score){
    print("You won!")
  } else if (you_score==computer_score){
    print("tie game")
  } else {
    print("Computer won!")
  }
}
  
#you=sample(c("rock","paper","scissors"), 1, replace=TRUE)
you=readline("please enter your action: rock, paper, scissors > ")
computer=sample(c("rock","paper","scissors"), 1, replace=TRUE)

score=rps(you,computer)
winner(score[1],score[2])

```


## version 5
Let's revisit the "rock, paper, scissors" game. After each round, the program will ask you whether you want to continue to play another round. The program ends if you enter "no"; otherwise, the program will continue to the next round.

This is also an iterative task, but we do not know in advance how many iteration will be. Thus, we can use while-loop.
```{r, eval=FALSE}
continue="yes"   # the logical expression in the while loop
total_score=c(0,0)

while (continue!="no"){
  
  you=readline("please enter your action: rock, paper, scissors > ")
  computer=sample(c("rock","paper","scissors"), 1, replace=TRUE)

  score=rps(you,computer)
  total_score=total_score+score
  
  continue=readline("Do you want to play another round (yes/no): ")
}

winner(total_score[1],total_score[2])

```

Congratulation! You have just created your first computer game in R!
