---
output:
  pdf_document: default
  html_document: default
---
# Visualization with ggplot2 - Animation

Animation is a great visual way to tell a story with data. There is a time dimension in your data, i.e., the data is observed at multiple time point, you can consider to do an animation to show how your data evolve over time.


First, we need to install the gganimate package for making annimation
```{r, eval=FALSE, message=FALSE}
install.packages("gganimate")
install.packages("gifski")
```

```{r, message=FALSE}
library(gganimate)
library(gifski)
library(data.table)
```


We will study how to build animation with ggplot2 and gganimate packages. We will use the gapminder dataset for illustration.
```{r, include=FALSE, eval=FALSE}
# Run the code below to generate the data.
install.packages("gapminder")
library(gapminder)
fwrite(gapminder,"data/gapminder.csv")
```

Read the gapminder.csv into R and examine the first few rows of the data using head().
```{r}
gapminder=fread("data/gapminder.csv")
head(gapminder)
```

The data shows the gdpDerap,lifeExpectation, population of different countries from 1952 to 2007.

##  Animation of scatter plot to show how two variables evolve over time simultaneously
Make a scatter between lifeExp and gdpDerap for all the years. Use the population to denote the size of the point.
```{r, include=FALSE, eval=FALSE}
p <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(show.legend = FALSE, alpha = 0.7) +
  scale_color_viridis_d() +
  scale_size(range = c(2, 12)) +
  scale_x_continuous(trans = "log10") +
  labs(x = "GDP per capita", y = "Life expectancy")
p
```

```{r}
p <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(show.legend = FALSE, alpha = 0.7) +
  scale_x_continuous(trans = "log10") +
  labs(x = "GDP per capita", y = "Life expectancy")
p
```

Create the gif to show each year a time:
```{r}
g<-p + transition_time(year) +
  labs(title = "Year: {frame_time}")

# output GIF
animate(g, renderer = gifski_renderer())

anim_save("data/gdp1.gif")
```

Facets by continent:
```{r}
p <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(show.legend = FALSE, alpha = 0.7) +
  scale_x_continuous(trans = "log10") +
  labs(x = "GDP per capita", y = "Life expectancy")

g<-p + facet_wrap(~continent) +
  transition_time(year) +
  labs(title = "Year: {frame_time}")

# render into gif
animate(g, renderer = gifski_renderer())

anim_save("data/gdp2.gif")
```


Add text to the plot
```{r}
text_dat=gapminder[country%in%c("United States", "China","United Kingdom","Japan")]

p=ggplot(gapminder) +  
  geom_point(aes(gdpPercap, lifeExp, size = pop, color = country), show.legend = FALSE, alpha = 0.7) +
  geom_text(data=text_dat, aes(gdpPercap, lifeExp, label=country), size=4, show.legend = FALSE) +
  scale_x_continuous(trans = "log10") +
  labs(x = "GDP per capita", y = "Life expectancy")
p 

g<-p + transition_time(year) +
  labs(title = "Year: {frame_time}")

# render into gif
h<-animate(g, renderer = gifski_renderer())
h

anim_save("data/gdp3.gif",h)
```
  
  
The animations shows a few observations:

* Overall all, gdpPercap and lifeExp increase across the world over time.

* It is clear that in Africa, a group of country is ahead of others.

* Europe countries are similar.

* US and UK have always among the top country in terms of GDP and lifeExp

* Japan is the most health country

* China has a large population and grows very fast in lifeExp and GDP.


## Use a line chart to show how gdpPercap evolve over time
Here we create line chart to show how gdpPercap evolve over time. We will then add animation to the chart to gradually show the line evolve over time.
```{r}
ggplot(gapminder, aes(year, gdpPercap, color = country)) +
  geom_line(show.legend = FALSE, alpha = 0.7) +
  labs(x = "Year", y = "GDP per capita", color="country")+
  theme_minimal() + 
  facet_wrap(~continent) 
```


It would different to show all countries in one single chart. Suppose we are particular intereted in the following country: "United States", "China","United Kingdom","Japan", "Korean".   
Use a line chart to show the gdpPercap of these countries evolve over time. 

```{r}
sub_gapminder=gapminder[country%in%c("United States", "China","United Kingdom","Japan","Korean")]

ggplot(sub_gapminder, aes(year, gdpPercap, color = country)) +  
  geom_line() +
  theme_minimal()+ 
  labs(x = "GDP per capita", y = "GDP Per Captia")
```

Let the line show gradually over time through animation:
```{r}
p=ggplot(sub_gapminder, aes(year, gdpPercap, color = country) ) +  
  geom_line(show.legend = FALSE) +
  geom_text(aes(year, gdpPercap, label=country), vjust = 0.2, hjust =0, size=4, show.legend = FALSE) +
  theme_minimal()+ 
  labs(x = "GDP per capita", y = "GDP Per Captia")

# create annimation   
g<-p + transition_reveal(year)

h<-animate(g, renderer = gifski_renderer())
h

anim_save("data/gdp4.gif",h)
```

Note that the value of hjust and vjust are only defined between 0 and 1: 0 means left-justified; 1 means right-justified; hjust controls horizontal justification and vjust controls vertical justification.


## Animation with Bar chart
We can make bar chart with animation to show the top n categories over time.
E.g., we want to use an animation to show the top 10 country in terms of gdpPercap. 

First, we need to subset the dataset to obtaint the the top 10 country in terms of gdpPercap for each year. This can be done using the powerful subsetting function in data.table package.
```{r}
# select the top 10 country for each year 
gapminder10 = gapminder[order(year,-gdpPercap)][, .SD[1:10],by = year]

# create rank of gdpPercap among country for each year
gapminder10[, rank:= .N:1, by = year]

head(gapminder10,15)
```

```{r}
p=ggplot(gapminder10) +
  geom_bar(aes(x=gdpPercap, y= factor(rank), fill = country), stat = "identity", show.legend = FALSE)+  
  geom_text(aes(x = -1000, y = factor(rank), label = country), vjust = 0.2, hjust = 1, size = 4,show.legend = FALSE) +
  labs(x = "GDP per capita", y = NULL) +
  scale_x_continuous(breaks = seq(0, 90000, 10000), expand = expansion(mult = c(.2, 0.05))) +
  theme(legend.position="none", 
        axis.text.y = element_blank(), 
        axis.title.y=element_blank(),
        axis.ticks.y=element_blank(),
        panel.background=element_blank())

g<-p + transition_time(year) +
  labs(title = "Year: {frame_time}")

# nframes: sets the total number of frames for the gif; 
# fps: frames per second to control the speed of gif.
h<-animate(g, nframes = 100, fps = 10, renderer = gifski_renderer())
h
anim_save("data/gdp5.gif",h)
```

Note that:

* Put the labels on the left of the datapoints by changing the horizontal alignment via hjust = 1.

* I also added some space between the label and the axis by setting x = -1000

* We have to add some space for the labels so that they are not cut when reaching the borders of the plot. This can be achieved by increasing the expansion of the scale via the expand argument of scale_x_continuous. I inreased the expansion at the lower end to 20 percent, while keeping the default 5 percent at the upper end.

* Finally, to prevent breaks with negative values when setting x = -1000 in geom_text I force the breaks to start at 0 via breaks = seq(0, 90000, 10000).


nframes is the number of frames. The greater the number, the better the transition. This is similar to drawing cartoons. However, with more frames, the processing time is longer. R will take some more time and consume more power.
fps is the amount of frame shown per second(default is 10)


Since year is a distinct value, we can also set the animation of transition with discrete state.
```{r}
g<-p + transition_states(year,transition_length = 3, state_length = 1) +
  labs(title = "Year: {closest_state}")

# output GIF
h<-animate(g, nframes = 100,fps = 15, renderer = gifski_renderer())
h
anim_save("data/gdp6.gif",h)
```

In above code,  

* transition_length: The relative length of the transition. 

* state_length: The relative length of the pause at the states. 






